---
title: "Languages spoken at home in the United States"
format: html
editor: visual
---

```{r}
#| include: false
library(tidyverse)

path_data <- "data/languages.csv"

# Color palette for plots
pal <- c(
  "#447099",
  "#EE6331",
  "#404041",
  "#419599",
  "#72994E",
  "#9A4665"
)
```

```{r}
#| message: false
languages <- read_csv(path_data) 
```

## Explore

`glimpse()`, from the dplyr package, provides a high-level overview of the data set.

```{r}
glimpse(languages)
```

The following code shows us the total number of at-home speakers of each language, sorted by the number of speakers. Page through the output to see the totals for other languages.

```{r}
languages |>
  group_by(language) |>
  summarize(total = sum(speakers, na.rm = TRUE)) |>
  arrange(desc(total))
```

Spanish has the most at-home speakers of any non-English language.

## Top languages spoken in MA

Spanish, Portuguese, and Chinese are the most common non-English languages spoken in Massachusetts. You can edit the code inside `filter()` to visualize a different state.

```{r}
languages |>
  filter(state == "Massachusetts") |> # Edit this line to explore a different state
  mutate(language = fct_reorder(language, percent, .desc = TRUE)) |>
  slice_max(percent, n = 9) |>
  ggplot(aes(language, percent)) +
  geom_col(fill = pal[[1]]) +
  theme_minimal()
```

How does Massachusetts compare to Connecticut?

```{r}
languages |>
  filter(state %in% c("Massachusetts", "Connecticut")) |> # Edit this line to compare other states
  mutate(language = fct_reorder(language, percent, .desc = TRUE)) |>
  slice_max(percent, n = 9) |>
  ggplot(aes(language, percent, fill = state)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = pal[1:2]) +
  theme_minimal() +
  labs(title = "Languages spoken at home in CT and MA")
```

### States where Tagalog is spoken most commonly

```{r}
languages |> 
  filter(language == "Tagalog") |> # Edit this line to explore a different language
  slice_max(order_by = percent, n = 10) |> 
  mutate(state = fct_reorder(state, percent, .desc = TRUE)) |> 
  ggplot(aes(state, percent)) +
  geom_col(fill = pal[[1]]) +
  theme_minimal() +
  labs(title = "Tagalog speakers by state")
```

## Most common language by state

In 47 states, Spanish is the most common non-English language spoken at home.

```{r}
languages |> 
  group_by(state) |> 
  filter(speakers == max(speakers)) |> 
  ungroup() |> 
  count(language, sort = TRUE) |> 
  rename(`Most common language` = language)
```

The only states where Spanish is not the most common language spoken are Alaska, Hawaii, Maine, and Vermont.

```{r}
languages |> 
  group_by(state) |> 
  filter(speakers == max(speakers)) |> 
  ungroup() |> 
  filter(language != "Spanish")
```

Two of these states (Alaska and Hawaii) states have multiple official languages:

-   Hawaii: English and Hawaiian (included in "Ilocano, Samoan, Hawaiian, or other Austronesian languages")

-   Alaska: English and 20 Alaska Native languages, including Yup'ik and Inupiaq (included in "Other Native languages of North America")

## Maine and Vermont

Maine and Vermont both have a relatively large proportion of French speakers. Other common languages include German, Spanish, and Chinese.

```{r}
languages |> 
  filter(
    state %in% c("Maine", "Vermont"),
    language %in% c("German", "Spanish", "Chinese", "French") 
  ) |> 
  ggplot(aes(language, percent, fill = state)) +
  geom_col() +
  scale_fill_manual(values = pal[1:2]) +
  facet_wrap(vars(state)) +
  guides(fill = "none") +
  theme_minimal()
```
